Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-11 aistats.tex
Update Optim section
Move MAP=SMD there for simplicity.
Compress the story by immediately introducing SMD, and skipping MD.

\usepackage[capitalise]{cleveref}
\newcommand{\rlp}[1]{\textcolor{red}{RLP:#1}}
\newcommand{\fdk}[1]{\textcolor{Periwinkle}{fdk:#1}}
\newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}
\newcommand{\tocite}{\textcolor{purple}{(add citation)}}

@@

@@

Exponential Families are to statistics what linear models are to regression.
Many classical model are exponential families, and their convexity and modeling flexibility make them one of the most widely used statistical models.
Here we review some of their properties.
We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
The exponential family for data $X\in \cX$ with natural (or primal) parameter $\nat$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
\begin{equation}
	 p(X|\nat) = \exp( \langle \nat, T(X) \rangle - \logpart(\nat)) \; ,
\end{equation}
where $\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.
Here $\logpart$ is the log-partition function -- i.e. the normalization term
\begin{align}
    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \; .
\end{align}
When explicit, we will write the random variable $T = T(X)$ for lightness.
For convenience, we focus on steep, regular exponential families
where $\logpart$ is strictly convex function of Legendre type,
and the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex
(See \citet{barndoffnielsen2014information}).
This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

@@

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat)
    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat))
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
Intuitively, $n_0$ is a number of fictive data points observed from a distribution with parameter $\nat_0$.

\paragraph{Maximum A Posteriori (MAP).}
Using Bayes rule, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
\begin{align}
	p(\nat \cond \mD_n)
	%\propto p(\mD|\nat)p(\nat)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt))
    \label{eq:joint_likelihood}
\end{align}
reaching its maximum in $\MAPt$, the MAP estimate verifying the first order optimality conditions
\begin{align}
    \nabla \logpart(\MAPt) = \MAPm
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \m_n^\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
\end{align}
This last equation is known as moment matching.
The MLE and MAP estimates are statistics of the dataset $\mD_n$.
Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.

@@

\subsection{MAP as Stochastic Mirror Descent}
\label{ssec:MAP=SMD}
By Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
Plugging in~\eqref{eq:joint_likelihood}, we see that the MAP verifies
\begin{align*}
	\hat \nat_{n}
	= \argmin_\nat - \log p(X_{n} | \nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
    %\label{eq:SBPP}
\end{align*}
which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$. Expanding $p(X_{n} | \nat)$ and  playing with the Bregman we get
\begin{align}
		\hat \nat_{n}
	= \argmin_\nat g_{n}^\top \nat + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
    \label{eq:SMD}
\end{align}
where $g_{n} := \nabla\logpart(\hat \nat_{n-1}) - T(X_n) = \hat \m_{n-1} - T_n$ is the gradient of $-\log p(X_{n} | \nat)$ evaluated at $\hat \nat_{n-1}$.
We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$ -- a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
This parallel means we could obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
In particular, it could potentially apply to a wide range of $\cD$.
Note that this optimization perspective does not hold for the MLE %($n_0=0$)
as the initial prior uniform prior may correspond to an initialization outside of $\Theta$.
We focus more on this approach in Section~\ref{sec:optimization}.

@@

\section{Optimization Perspective}
\label{sec:optimization}
As we saw in Section~\ref{ssec:MAP=SMD}, MAP can be interpreted as Stochastic Mirror Descent (SMD).
There are two consequences of this observation: 1. we may obtain a convergence rate for MAP from an optimization analysis, and 2. any insights gained from the MAP may inform further design and analysis of SMD.

Mirror Descent (MD)\footnote{%
Also known as is going by many names:
Bregman (proximal) Gradient, Relative Gradient Descent, NoLips.
}
\citep{nemirovski1983problem}
was first introduced to optimize non-smooth functions \fdk{(Is typically encountered in non-smooth, online optimization?)},
typically under bounded gradient assumption on the objective $f$
or strong convexity assumption on the potential $\logpart$

\fdk{
{\bf Online learning paragraph?}
\citet{azoury2001relative,freund1996predicting,dasgupta2007online},
Bubeck's paragraph
}

More recently, advances in optimization that go beyond the Euclidean case,
to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
neither strongly convex in the Euclidean norm, using the concept of relative smoothness
\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.
Instead of measuring strong-convexity and smoothness in the Euclidean norm (or any norm),
the $\mu$-strong convexity and $L$-smoothness of a function $f$
relative to a reference function $A$ is defined  as
\aligns{
	\mu \cB_{A}(x, y)
	\leq
	\cB_f(x,y)
	\leq
	L \cB_A(x,y).
}
Those conditions ensure the linear convergence rate of MD with $A$ as the reference function.
Exponential families log-likelihood perfectly fits into this theory.
\aligns{
	f(\theta) = A(\theta) - \expect[x\sim p(x\cond\theta)]{\lin{x, \theta}}
}
is $1$-smooth and $1$-strongly convex relative to $A$.
However, the MAP is stochastic, and known results for stochastic mirror descent do not give convergence rates
for all exponential families.

\subsection{Strongly-convex reference function}

{Traditional results}
in stochastic optimization,
such as the work of \citet{nemirovski2009robust,ghadimi2012optimal},
rely on the assumption that the reference function $A$
is itself strongly convex relative to a norm $\norm{\cdot}$,
(meaning that $f$ is strongly convex).
This apply for the MLE problem of a Gaussian distribution
with fixed variance,
but not in a general case.
when the covariance is unknown,
or for a Poisson distribution, as $-\log(\theta)$ and $\exp(\theta)$
are not strongly convex.

The introduction of relative smoothness has renewed interest for analyses of mirror descent
without relying on strong-convexity in an underlying norm.
Recent works have made attempts to generalize the assumptions
that are used to control the convergence of stochastic gradient descent
to this setting.
Those include
bounded gradients $(\expecti{\norm{\tilde g(\theta)}^2} \leq C$,
bounded variance $(\expecti{\norm{\nabla f(\theta) - \tilde g(\theta)}^2} \leq C$
uniformly or only at the minimum $\theta_*$,
or bounded disparities between the minimum of $f$ at $\theta_*$ and the minima of individual functions $f_i$,
if $f(\theta) = \expect[i]{f_i(\theta)}$ $(\expect{f(\theta_*) - \min_\theta f_i(\theta)} < C)$.
(\fdk{For a review of those assumptions, see that paper by Richtarik's students})

However, they do not yet address the problem satisfactorily.
We present those generalization here
and they are summarized in the context of maximum likelihood estimation
in \cref{tbl:assumptions}.

@@

\subsection{Bounded variance}
An adaptation of the bounded variance,
investigated by \citet{hanzely2018fastest},
defines the following bounded quantity.
Letting $\theta^+(v)$ be a step from $\theta$ with vector $v$,
$\theta^+(v) = \nabla A^*(\nabla A(\theta) - v)$,
\alignn{\label{eq:assumption-hanzely}
	\expect[\tilde g\!]{
	\lin{
		\nabla\! f(\theta)
		- \tilde g(\theta),
		\theta^+\!\!\paren{\eta \tilde g(\theta)}
		- \theta^+\!\!\paren{\eta \nabla\! f(\theta)}
	}
	}
	\leq
	\eta C,
}
for some step-size $\eta$. Let us unpack this quantity;

{\bf For a Gaussian} with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$),
this definition recovers the variance of the stochastic gradient
$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C$.
More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
the LHS of \cref{eq:assumption-hanzely} is bounded by
$\expecti{\frac{1}{\mu}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$,
and a bound on the variance of the stochastic gradient is a sufficient condition for \cref{eq:assumption-hanzely}
to hold.

{\bf In divergence,} this condition states that the (symmetrized) divergence between
a deterministic and stochastic update is bounded in expectation,
\alignn{\label{eq:hanzely-in-divergence}
	\begin{aligned}
	&
	\mathbb{E}_{\tilde g}\bigg[
	\cB_{A^*}\paren{\m - \eta \tilde g(\theta), \m - \eta \nabla f(\theta)}
	\\[-1em]
	&\hphantom{\mathbb{E}_{\tilde g}[}
	+ \cB_{A^*}\paren{\m - \eta \nabla f(\theta), \m - \eta \tilde g(\theta)}
	\bigg]
	\leq
	\eta C.
	\end{aligned}
}
{\bf For a maximum likelihood problem, }
where $\tilde g(\theta) = \m - x$
where $x \sim p(x\cond\theta_*)$,
and $\nabla f(\theta) = \m - \m_*$,
\aligns{
	&
	\mathbb{E}_{x}\bigg[
	\cB_{A^*}\paren{(1-\eta) \m + \eta x, (1-\eta)\m + \eta \m_*}
	\\
	&\hphantom{\mathbb{E}_{\tilde g_t}[}
	\cB_{A^*}\paren{(1-\eta)\m + \eta \m_*, (1-\eta) \m + \eta x}
	\bigg]
	\leq
	\eta C.
}
\fdk{Could work with decreasing step-sizes}

\fdk{How do we show that this fails?}

\subsection{Bounded variance at the minimum}

@@

\textbf{Bounded optimality gap,}
\fdk{
Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}?
\\
Do they do something with that? It seems to be fixed step-size only.
}
\aligns{
	A(\theta_*) - \lin{x, \theta_*} - \expecti{ \min_\theta A(\theta) - \lin{x, \theta} }
}

>>>>

\usepackage[capitalise]{cleveref}
\newcommand{\rlp}[1]{\textcolor{BrickRed}{(RLP:#1)}}
\newcommand{\fdk}[1]{\textcolor{Periwinkle}{(fdk:#1)}}
\newcommand{\TODO}[1]{\textcolor{cyan}{(TODO #1)}}
\newcommand{\tocite}{\textcolor{purple}{(add citation)}}

@@

\paragraph{Notation}
$\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.

@@

Exponential Families are to statistics what linear models are to regression.
Many classical model are exponential families, and their convexity and modeling flexibility make them one of the most widely used statistical models.
Here we review some of their properties.
We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
The exponential family for data $X\in \cX$ with natural (or primal) parameter $\nat$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
\begin{equation}
	 p(X|\nat) = \exp( \langle \nat, T(X) \rangle - \logpart(\nat)) \; .
	 \label{eq:def_expfamily}
\end{equation}
Here $\logpart$ is the log-partition function -- i.e. the normalization term
\begin{align}
    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \; .
\end{align}
When explicit, we will write the random variable $T = T(X)$ for lightness.
For convenience, we focus on steep, regular exponential families
where $\logpart$ is strictly convex function of Legendre type,
and the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex
(See \citet{barndoffnielsen2014information}).
This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

@@

\paragraph{Conjugate Prior}
One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
\begin{align}
    p(\nat)
    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat))
    \label{eq:def_prior}
\end{align}
where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
Intuitively, $n_0$ is a number of fictive data points observed from a distribution with parameter $\nat_0$.

\paragraph{Maximum A Posteriori (MAP).}
Using Bayes rule with~\eqref{eq:def_expfamily}~\&~\eqref{eq:def_prior}, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
\begin{align}
	p(\nat \cond \mD_n)
	%\propto p(\mD|\nat)p(\nat)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
reaching its maximum in $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \m_n^\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
\end{align}
This last equation is known as moment matching.
In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.
The MLE and MAP estimates are statistics of the dataset $\mD_n$.
Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.

@@

@@

\section{Optimization Perspective}
\label{sec:optimization}

We now show how MAP can be interpreted as stochastic mirror descent (SMD).
This means that 1. we may obtain a convergence rate for MAP from an optimization analysis, and 2. any insights gained from the MAP may inform further design and analysis of SMD.
We then review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating 3 analysis of SMD with the MAP.
\rlp{mention that examples have so far been lacking for relative SMD.}

\subsection{MAP as Stochastic Mirror Descent}
\label{ssec:MAP=SMD}
By Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
Plugging in~\eqref{eq:joint_likelihood}, and introducing $f_n(\nat) = - \log p(X_{n} | \nat)$, we see that the MAP verifies
\begin{align*}
	\hat \nat_{n}
	= \argmin_\nat f_n(\nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
    %\label{eq:SBPP}
\end{align*}
which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$.
%Maybe skip this SBPP story altogether
Expanding $f_n(\nat) = \logpart(\nat) - T_n$ and  playing with the Bregman we get
\begin{align*}
		\hat \nat_{n}
	= \argmin_\nat \langle g_{n}(\hat \nat_{n-1}), \nat \rangle + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
    %\label{eq:SMD}
\end{align*}
where $g_{n}(\nat) := \nabla f_n(\nat) = \hat \m_{n-1} - T_n$.
We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$.
From the first order optimality condition, we see this is a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
This parallel means we could obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
In particular, it could potentially apply to a wide range of distribution $\cD$ on $X$.
Note that this optimization perspective does not hold for the MLE %($n_0=0$)
as its implicit uniform prior may correspond to an initialization outside of $\Theta$.

\subsection{Relative Smoothness}
Mirror descent (MD)\footnote{
Also known as
Bregman (proximal) gradient, relative gradient descent or NoLips.
}
\citep{nemirovski1983problem,beck2003mirror}
%Nemirovski introduced the algorithm, and Beck established the connection with Bregman divergences, e.g. the argmin form.
and SMD 
\citet{nemirovski2009robust,ghadimi2012optimal}  \TODO{verify what these citations are talking about} 
are typically encountered in non-smooth (online) optimization,
under bounded (or Lipschitz) gradient assumption on the objective $f$
and strong convexity assumption on the potential $\logpart$
\citep[Th.4.2 \& Th.6.3]{bubeck2015convex}.
In our case, none of these assumptions may hold. 
For instance $\logpart = -\log$ is neither smooth nor strongly convex.

\fdk{
{\bf Online learning paragraph?}
\citet{azoury2001relative,freund1996predicting,dasgupta2007online},
Bubeck's 2011.
}

Recently, these assumptions have been relaxed to allow for non-smooth functions $f$
\footnote{$f$ is $L$-smooth if its gradient is $L$-Lipschitz}.
and non-strongly convex potential $\logpart$, first in the deterministic setting
\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}, then in the stochastic setting \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.
These assumptions have been replaced with
the $\mu$-strong convexity and $L$-smoothness of $f$
\emph{relative} to $\logpart$, defined as
\aligns{
	\mu \cB_{A}(x, y)
	\leq
	\cB_f(x,y)
	\leq
	L \cB_A(x,y) \; .
}
When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent. 
Those conditions ensure the linear convergence rate of MD with $A$ as the reference function.
Exponential families log-likelihood perfectly fit into this framework, as
\aligns{
	f(\theta) = A(\theta) - \expect{\lin{X, \theta}}
}
is $1$-smooth and $1$-strongly convex relative to $A$.

\subsection{Bounding the Randomness}
Smoothness is not the only issue.
To analyze stochastic algorithms, one also need to quantify the randomness. 
We now review these modern attempts at analyzing SMD, and illustrate their limits  in the context of the MAP.
See \cref{tbl:assumptions} for a summary.

@@

\subsubsection{Bounded variance}
\citet{hanzely2018fastest} assume the covariance between negative gradient and next iterate is upper bounded at every time step
\alignn{
\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C, \forall n \geq 1
\label{eq:assumption-hanzely}
}
for some constant $C$.
For a Gaussian with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$),
this definition recovers the variance of the stochastic gradient
$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C$.
More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
the LHS of \cref{eq:assumption-hanzely} is bounded by
$\expecti{\frac{1}{\mu}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$,
and a bound on the variance of the stochastic gradient is a sufficient condition for \cref{eq:assumption-hanzely}
to hold.

Under this assumption, they prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size \citep[Lem.4.8]{hanzely2018fastest} , which is precisely what we wish to obtain.

In general however, this assumption hardly holds.
For the MAP, this translates into
\alignn{
\expect[T_n]{\lin{T_n - \mu^* , \MAPt}} \leq \lr_n C, \forall n \geq 1
}
which fails to hold on the gaussian variance example for instance. \TODO{short proof or formalize.}

\subsubsection{Bounded variance at the minimum}

@@

\subsubsection{Bounded optimality gap,}
\fdk{
Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}?
\\
Do they do something with that? It seems to be fixed step-size only.
}
\aligns{
	A(\theta_*) - \lin{x, \theta_*} - \expecti{ \min_\theta A(\theta) - \lin{x, \theta} }
}
